---
title: 'Homework: Finite Precision Arithmetic & Gradient Descent'
output:
  html_document:
    df_print: paged
  html_notebook:
    code_folding: none
    highlight: textmate
---


```{r setup, include=FALSE}
source(file.path("..", "R", "colors.R"))
source(file.path("..", "R", "util.R"))

required_packages <- c('Matrix', 'testthat')
install_and_load_packages(required_packages)
```

# Exercise 0: Checking your understanding of floating point numbers and finite precision arithmetic

We observed in the class that the following code returns `FALSE`:

```{r, eval=FALSE}
a <- 1 + .Machine$double.eps
b <- 1 + .Machine$double.eps 
c <- 1
(a + b) + c == a + (b + c)
```

On the other hand, the following modification results in `TRUE`:

```{r, eval=FALSE}
b <- 0.5 + .Machine$double.eps
(a + b) + c == a + (b + c)
```

Why do we get the different result?

**Answer:**

There're \( 2^{52} \) machine-representable values in between two reals \( 2^{k} \) and \( 2^{k+1} \) w/ \( 2^{k} \times 2^{-52} \) increment. 

Notice, for numbers between \( 2^{1} \) and \( 2^{2} \), the smallest increment is \( 2^{-52} \times 2^{1} = 2^{-51} \), and for numbers between \( 2^{0} \) and \( 2^{1} \), the smallest increment is \( 2^{-52} \times 2^{0} = 2^{-52} \).

For the first equation, b+c is within the range of \( 2^{1} \) and \( 2^{2} \), thus the epsilon is not distinguishable in \( b + c = 2 + \epsilon \) and it becomes \( b + c = 2 \). And eventually \( a + (b + c) = 2 + 1 = 3 \) and \( (a + b) + c = 2 + 1 = 3 + 2 \epsilon \), thus they are not the same.

However, for the second equation, b+c is within the range of \( 2^{0} \) and \( 2^{1} \), thus the epsilon is distinguishable in \( b + c = 1.5 + \epsilon \). So the equations get the same results, both eaqual to \( 2.5 + \epsilon \).


# Exercise 1: One-pass algorithm for calculating variance

Let's say you are running a Monte Carlo simulation to estimate the mean and variance of some random variable $\boldsymbol{\theta}$. 
Imagine that $\boldsymbol{\theta}$ is high-dimensional, so you don't want to keep all the Monte Carlo samples $\boldsymbol{\theta}_i$ due to the memory burden.
(If $\boldsymbol{\theta}$ is $10^6$-dimensional, for example, storing 1,000 Monte Carlo samples would require 8GB of memory.)
And it occurs to you that you can write a "one-pass" algorithm to calculate the sample mean and sample second moment of $\boldsymbol{\theta}$ by only keeping track of running totals of Monte Carlo samples. 
You can then, in theory, calculate the sample variance from these quantities.

The above idea is implemented and applied below. 
Try it and see what you get. 
Does the estimate make sense? If not, why?
How does the estimate change if you change the seed?

**Bonus question:** 
What is, roughly speaking, the distribution of the estimator `sample_var_with_question_mark` under repeated Monte Carlo simulations (i.e. without a fixed seed)?

```{r}
set.seed(140778)

running_sum <- 0
running_sum_sq <- 0

n_sample <- 10^3
for (i in 1:n_sample) {
  theta <- 1e10 + rnorm(1)
  running_sum <- running_sum + theta
  running_sum_sq <- running_sum_sq + theta^2
}
sample_mean <- running_sum / n_sample
sample_sec_mom <- running_sum_sq / n_sample

sample_var_with_question_mark <- (sample_sec_mom - (sample_mean)^2)
```

**Answers:**

The estimate for mean makes sense but the variance does not. The variance should be close to 1, but the estimate is way off from 1, and even negative. Because the correct way for estimating variance is to subtract the sample mean from each sample, and then square and sum them up, and then divide by n-1. Which is clearly not equivalent to square the samples and sum them up and divide by n, then subtract the mean's square.

When I change the seed, the estimate changes a lot, from -147456 to -81920.


**Bonus answer:**
```{r}
res <- c()
for(it in 1:1000){
  running_sum <- 0
  running_sum_sq <- 0
  
  n_sample <- 10 ^ 3
  for (i in 1:n_sample) {
    theta <- 1e10 + rnorm(1)
    running_sum <- running_sum + theta
    running_sum_sq <- running_sum_sq + theta ^ 2
  }
  sample_mean <- running_sum / n_sample
  sample_sec_mom <- running_sum_sq / n_sample
  
  res <- c(res,sample_var_with_question_mark <- (sample_sec_mom - (sample_mean) ^ 2))
}
hist(res)
```

Turns out to be a normal distribution. This is due to central limit theorem since the random variable (sample_sec_mom) is the sum of many iid random variables (in this case a shifted chi-square distribution) abd the sum should approaching a normal distribution, and (sample_mean) ^ 2 has really small variance thus almost constant. Thus the difference between them is approximately a normal distribution.

## Things to commit/submit
Answers to the questions above.


# Exercise 2: Generating random real numbers in finite precision arithmetic

## Part A

You learned in the class that numeric values in computers are typically represented with 64-bits double-precision numbers, having 52-bits fraction. 
In particular, a "Uniform(1, 2)" random variable in double-precision generates a number from the $2^{52}$ unique values between $1$ and $2$.
The amount of randomness in it, therefore, is equivalent to 52 Bernoulli random variables.
Write a function to generate a Uniform(1, 2) random variable in double-precision with randomness coming only from the `rbinom()` function.
Check the function's behavior by plotting a histogram of random variables generated from it. 

```{r, eval=TRUE}
double_prec_runif <- function(){
  # Fill in
  decimals <- 1/2^(1:52)
  digits <- rbinom(52, size = 1, prob = 0.5)
  return(decimals %*% digits + 1)
}

n_draw <- 10^5
unif_rv <- sapply(1:n_draw, function (i) double_prec_runif())

rv_hist <- hist(unif_rv, breaks = 21, plot = FALSE)
plot(
  rv_hist,
  freq = FALSE,
  xlab = "Realized values",
  col = "#bdcedb", # JHU Spirit Blue with saturation and luminance adjusted: https://hslpicker.com/#bdcedb
  border = "#859bad" # Another variation of JHU Spirit Blue
)
```

## Part B

As we observed above, a Uniform(1, 2) random variable in double-precision takes one of the $2^{52}$ unique values.
Two independently drawn random variables, therefore, has small but positive probability of having exactly the same value.
We can calculate the expected number of duplicate values in i.i.d. Uniform(1, 2) random variables as follows.
When drawing with replacement $d$ independent random variables out of $n$ choices, the probability of the $i$-th element never selected is $(1 - n^{-1})^d$ for $i = 1, \ldots, n$. 
The expected number of unique items drawn, therefore, is $n [1 - (1 - n^{-1})^d]$.
In turn, the expected number of duplicates is $d - n [1 - (1 - n^{-1})^d]$.
For the case of double-precision uniform random variables, we have $n = 2^{52}$ for the number of choices.

The `calc_num_expected_duplicate()` function below implements the above formula for the expected number of duplicates. 
When you call `calc_num_expected_duplicate(n_draw = 10^6, n_choice = 2^52)` to calculate the expected number for $10^6$ Uniform(1, 2) random variables, however, the function returns _exact zero_ due to a finite precision error. 
What is the issue?

```{r}
calc_num_expected_duplicate <- function (n_draw, n_choice) {
  # To be remedied.
  # prob_item_never_chosen <- (1 - 1 / n_choice) ^ n_draw
  # n_expected_unique <- n_choice * (1 - prob_item_never_chosen)
  d <- n_draw
  n <- n_choice
  res <- d*(d-1)/(2*n)
  return(res)
}
calc_num_expected_duplicate(n_draw = 10^6, n_choice = 2^52)
```

**Issue:**

n_choice * (1 - prob_item_never_chosen) is indistinguishable from n_choice (1e+06) due to finite precision arithmetic (1e+06 is between \( 2^{23} \) and \( 2^{24} \), with a minimum distinguishable increment of 199840144), thus the returned result will be 1e+06 - 1e+06 = 0.


Using second order taylor expansion, we can simplify the formula $d - n [1 - (1 - n^{-1})^d]$

Firstly, $(1-x)^{d} \approx 1 - xd + \frac{1}{2}x^{2}d(d-1)$, so $(1 - n^{-1})^d \approx 1 - d n^{-1} + \frac{1}{2}d(d-1)n^{-2}$.

Then, $d - n [1 - (1 - n^{-1})^d] \approx d - n [1 - (1 - d n^{-1} + \frac{1}{2}d(d-1)n^{-2})] = d - n(dn^{-1}-\frac{1}{2}d(d-1)n^{-2}) = d - d + \frac{1}{2n}d(d-1) = \frac{d(d-1)}{2n}$.

Formulas involving $1 + x$ for $|x| \ll 1$ ($x = - n^{-1}$ in the above case) occur commonly enough in statistics/scientific computing that R provides a function `log1p(x)` for computing $\log(1 + x)$ to deal with one such situation.
In more general calculations involving $|x| \ll 1$ (e.g. for a "sinc" funciton $\operatorname{sinc}(x) = \sin(x) / x$), a Taylor expansion provides a general approach to control truncation errors.

Remedy the `calc_num_expected_duplicate()` function by applying the second-order Taylor expansion to $(1 - x)^d$ and simplifying the formula $d - x^{-1} [1 - (1 - x)^d]$ to avoid truncation errors. 
(Remark: i) in this specific case, you could use the binomial expansion and calculate it exactly; 
ii) applying the second-order expansion to $(1 - x)^d$ corresponds to applying the first-order expansion to the overall expression $d - x^{-1} [1 - (1 - x)^d]$.)
Use the more numerically accurate version of `calc_num_expected_duplicate()` to calculate the expected number of duplicates among $10^6$ Uniform(1, 2) random variables.

## Part C

The expected number from Part B, if you calculated it correctly, suggests that it is extremely rare to get duplicate values when generating $10^6$ uniform random variables in double-precision.
This is reassuring because we routinely generate a far larger number of random numbers in modern applications.
If we use R's `runif()` function, however, this is _not_ what we observe because [R's random numbers are generated using 32-bit integers](https://www.tandfonline.com/doi/full/10.1080/00031305.2020.1782261).
Check the number of duplicates from `runif(10^6)` against the expected value for a floating-point number with 32-bits of precision. 
Check also that, using your implementation of random uniform from Part A, duplicates are indeed rare in double-precision.

```{r, eval=TRUE}
n_draw <- 10^6
base_r_unif <- runif(n_draw, min = 1, max = 2)
unif_rv <- sapply(1:n_draw, function (i) double_prec_runif())
# Check the number of duplicates against the expected value
```
**Expected number of duplicates for 32 bits:**
```{r}
calc_num_expected_duplicate(n_draw, 2^32)
```

**Number of duplicates for base R uniform:**
```{r}
n_draw - length(unique(base_r_unif))
```

**Number of duplicates for double precision uniform:**
```{r}
n_draw - length(unique(unif_rv))
```


## Things to commit/submit
Answers to the tasks/questions above.


# Exercise 3: Numerical differentiation

## Part A: Potential caveats of numerical differentiation

You have learned in the class that the centered difference $\frac{f(x + \Delta x) - f(x - \Delta x)}{2 \Delta x}$ approximates the derivative $f'(x)$ up to an $O(\Delta x^2)$ error.
To numerically verify this, we design the following experiment using a log of logistic function (i.e. $f(x) = \log\{ \exp(x) / (1 + \exp(x)) \}$) as an example.
We evaluate its derivative using an analytical formula and compare it to the centered difference approximations as $\Delta x$ varies from $2^{-1}$ to $2^{-52}$.
Using the starter code provided below, plot the relative error of the numerical approximation in a log-log scale.
What do you expect the slope of the plotted line to be?

```{r, eval=TRUE}
#' Approximate derivative via centered-difference approximation
approximate_deriv <- function(func, x, dx) {
  # Fill in
  (func(x + dx) - func(x - dx)) / (2 * dx)
}

log_logistic_func <- function(x) {
  # Fill in
  log(exp(x) / (1 + exp(x)))
}

log_logistic_deriv <- function(x) {
  # Fill in
  1/(1 + exp(x))
}

# Calculate and plot the relative errors of finite difference approximations
set.seed(615)
x <- rnorm(1)

log2_dx <- - seq(1, 52)
numerical_deriv <- sapply(
  2^log2_dx, function(dx) approximate_deriv(log_logistic_func, x, dx)
)
analytical_deriv <- log_logistic_deriv(x)
rel_err <- abs(
  (analytical_deriv - numerical_deriv) / analytical_deriv
)

fontsize <- 1.2
plot(
  log2_dx, log2(rel_err),
  frame = FALSE, # Remove the ugly box
  col = jhu_color$spiritBlue,
  xlab = "log2 stepsize for finite difference",
  ylab = "log2 rel errors",
  xlim = rev(range(log2_dx)),
  cex.lab = fontsize,
  cex.axis = fontsize,
  type = 'l'
)
```

**Questions:**
Does the relative error plot agree with what the mathematical theory predicts? Explain the observed phenomenon.

Also, what is the slope in the "latter part" of your plotted line, roughly speaking? 
Explain why we observe such a slope in the relative error. 
(Hint: Consider a machine with 10 decimal digits of accuracy. What happens if we numerically differentiate $f(x) = x$ at $x = 1$ with $\Delta x = 0.1, 0.01, 0.001, \ldots$.)

**Answers:**
No, the first part of relative error plot where the line decreases does agree with what the mathematical theory predicts. The relative error should be decreasing as the step size decreases, since the error is on a $O(\Delta x^2)$ scale and if you differentiate against $- \Delta x$ you will get $-2O(\Delta x)$, that is why we are getting a decreasing line for the first part relative error plot where the line decreases.

but the latter part of the plot shows that the relative error is increasing as step size gets smaller and smaller, which contradicts to the math theory. Further more, the slope of the relative error is roughly 1. This is due to the finite precision arithmetic and the consequent rounding error. When the step size is too small, the difference between \( f(x + \Delta x) \) and \( f(x - \Delta x) \) is too small and almost remain constant due to the round off error, however the denominator \( 2 \Delta x \) is getting smaller and smaller (changing by $2^{-1}$ each time), thus the relative error is increasing and the slope is roughly 1, since $\epsilon_{rel} \approx 1/\Delta x$ and $log(\epsilon_{rel}) \approx -log(\Delta x)$ and if we change to y and x notation it will be $y = -x$, so differentiating against $-x$ will give us a slope of 1.

## Part B: Extention to multivariate functions

As mentioned in the lecture, the finite difference method extends to a function with multivariate input $f: \mathbb{R}^d \to \mathbb{R}$ and can be used to approximate the gradient $\nabla f(\boldsymbol{x})$.
The extension relies on the fact $f(\boldsymbol{x} + \Delta x \, \boldsymbol{e}_i) = f(\boldsymbol{x}) + \Delta x \, \partial_i f(\boldsymbol{x}) + \ldots$ for $i = 1, \ldots, d$.

Using the starter code provided below, implement a function to approximate $\nabla f(\boldsymbol{x})$ via the centered difference method.
To test this function, take as an example a quadratic function $f(\boldsymbol{x}) = - \boldsymbol{x}^\intercal \boldsymbol{\Phi} \boldsymbol{x} / 2$ with positive definite $\boldsymbol{\Phi}$, which can be thought of as a log-likelihood function for $\boldsymbol{x} \sim \mathscr{N}(\boldsymbol{0}, \boldsymbol{\Phi})$ up to an additive constant.
First show that its gradient is given by $\nabla f(\boldsymbol{x}) = - \boldsymbol{\Phi} \boldsymbol{x}$; 
more generally, the gradient of a quadratic function $g(\boldsymbol{x}) = \boldsymbol{x}^\intercal \boldsymbol{A} \boldsymbol{x}$ is given by $\nabla g(\boldsymbol{x} = (\boldsymbol{A} + \boldsymbol{A}^\intercal) \boldsymbol{x}$ for any square matrix $\boldsymbol{A}$.
Then compare a numerically approximated gradient to one computed via an analytical formula.

**Answers:**
The gradient of a quadratic function :

Given the function 

\[ g(\boldsymbol{x}) = \boldsymbol{x}^\intercal \boldsymbol{A} \boldsymbol{x} \]

we want to find its gradient \( \nabla g(\boldsymbol{x}) \).

The function \( g(\boldsymbol{x}) \) can be expanded into its scalar components:

\[ g(\boldsymbol{x}) = \sum_{i=1}^n \sum_{j=1}^n x_i A_{ij} x_j \]

Now take derivative:

\[ \frac{\partial g(\boldsymbol{x})}{\partial x_k} = \frac{\partial}{\partial x_k} \sum_{i=1}^n \sum_{j=1}^n x_i A_{ij} x_j \]

When differentiating, we treat \( x_k \) as the variable and all other \( x_i \)'s as constants. There are two cases to consider:

1. When \( i = k \), the derivative of \( x_i A_{ij} x_j \) with respect to \( x_k \) is \( A_{kj} x_j \) since \( A_{ij} x_j \) is treated as a constant coefficient.
2. When \( j = k \), the derivative of \( x_i A_{ij} x_j \) with respect to \( x_k \) is \( x_i A_{ik} \), by the same logic.

Combining both cases, we get:

\[ \frac{\partial g(\boldsymbol{x})}{\partial x_k} = \sum_{j=1}^n A_{kj} x_j + \sum_{i=1}^n x_i A_{ik} \]

The first term is the \( k \)-th component of the matrix-vector product \( \boldsymbol{A} \boldsymbol{x} \), and the second term is the \( k \)-th component of the matrix-vector product \( \boldsymbol{A}^\intercal \boldsymbol{x} \) (since transposing swaps the indices).

Therefore, the gradient \( \nabla g(\boldsymbol{x}) \) is:

\[ \nabla g(\boldsymbol{x}) = \boldsymbol{A} \boldsymbol{x} + \boldsymbol{A}^\intercal \boldsymbol{x} \]

If \( \boldsymbol{A} \) is symmetric (\( \Phi \)), then this simplifies to \( 2 \Phi \boldsymbol{x} \). So we have \(\nabla f(\boldsymbol{x}) = - 2\boldsymbol{\Phi} \boldsymbol{x}/2 = - \boldsymbol{\Phi} \boldsymbol{x} \)


```{r, eval=TRUE}
approx_grad <- function(func, x, dx = .Machine$double.eps^(1/3)) {
  numerical_grad <- rep(0, length(x))
  D <- length(x)
  # Fill in
  for(d in 1:D){
    e_d <- rep(0, D)
    e_d[d] <- 1
    numerical_grad[d] <- (func(x + dx * e_d) - func(x - dx * e_d)) / (2 * dx)
  }
  return(numerical_grad)
}

set.seed(410)
n_param <- 4
X <- matrix(rnorm(2 * n_param^2), nrow = 2 * n_param, ncol = n_param)
Sigma_inv <- t(X) %*% X
  
#' Calculate log-density of centered Gaussian up to an additive factor
gaussian_logp <- function(x, Sigma_inv) {
  logp <- - .5 * t(x) %*% Sigma_inv %*% x
  return(logp)
}

gaussian_grad <- function(x, Sigma_inv) {
  grad <- -1 * Sigma_inv %*% x
  return(grad)
}

x <- c(3, 1, 4, 1)

analytical_grad <- gaussian_grad(x, Sigma_inv)
numerical_grad <- approx_grad(function(x) gaussian_logp(x, Sigma_inv), x)
testthat::expect_true(are_all_close(
  analytical_grad, numerical_grad, abs_tol = Inf, rel_tol = 1e-3
))
```


## Things to commit/submit
For Part A, completed code and answers to the questions.
For Part B, derivation of the gradient and completed code that passes the test.


# Exercise 4: Continuing development of `hiperglm` &mdash; finding MLE via `stats::optim()` and testing it

We will continue the development of the `hiperglm` package. 

As discussed, our plan is to first implement an MLE finder via a pre-packaged general-purpose BFGS optimizer, against which we can then use to benchmark our custom implementation of iteratively reweighted least squares. 
The idea is that using the pre-packaged function is less error-prone, especially since the gradient function is easy to unit-test.

To be extra safe, we are going to test the MLE finder via BFGS against an even simpler algorithm: the least square for a linear model via the analytical expression $\hat{\boldsymbol{\beta}}_\textrm{mle} = (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \boldsymbol{y}$.
Complete the following steps toward this goal.
Remember that you can use [`devtools::load_all()`](https://r-pkgs.org/code.html#sec-code-load-all) to help your [package development process](https://r-pkgs.org/code.html#constant-health-checks). 

0. Make and check out (`git branch` and `git checkout`) a development branch (call it `develop`) where you will commit the work below.
1. Set up the use of `testthat`.
2. Copy the helper functions `are_all_close()` and `simulate_data()` from the instructor's repo.
3. Test the `are_all_close()` function. 
Write a test to cover at least three cases: the function 1) correctly returns `TRUE`, 2) correctly returns `FALSE` because the relative error is above `rel_tol`, and 3) correctly returns `FALSE` because the absolute error is above `abs_tol`. 
Make sure that the logic of the test is clear and easy to understand &mdash; tests are there to clarify and validate the code's behavior; unreadable tests defeat the purpose.
<br> _Remark:_ This is just an exercise, but in general there is nothing strange about testing a function you use to test another function. It is perfectly reasonable to consider testing any functions complicated enough to potentially contain insiduous bugs.
4. Set up the (currently failing) test to compare the MLE estimated via pseudo-inverse and via BFGS. 
5. Implement the MLE finder via peudo-inverse. 
Please, pretty please, do not calculate an inverse of a matrix. 
(Otherwise, it's fine to use `solve()` instead of `chol()`.)
6. Implement the MLE finder via BFGS:
    a. Implement functions to calculate the log-likelihood and gradient under a linear model. 
    This function technically requires an estimate of the residual variance, but its value does not affect the MLE for the coefficients and hence is irrelevant for our purpose. 
    You can just specify it as an optional parameter e.g. as `function(..., noise_var = 1)`.
    b. Test your gradient calculation by comparing it against a numerical one.
    For this purpose, use the function from Exercise 2, Part B to approximate the gradient of a given function via finite difference.
    <br> _Remark:_ Technically, this test checks only for consistency between the log-likelihood and gradient function.
    It is possible to have both functions wrong but consistent with each other so that the test passes.
    This test nonetheless can catch many bugs and, since the gradient calculation tends to be more error prone, it's good to make sure there is no mistake there. 
    c. Plug in the now tested-and-true log-likelihood and gradient functions to `stats::optim()` to find MLE.
7. Run `devtools::test()` and check that the then-failing test from Step 2 now passes.
8. Open a pull request from your development branch to the main/master one. 
    
## Things to commit/submit

[Link to the pull request](https://github.com/thepianistalex/hiperglm/pull/1).
