---
title: 'Homework: Finite Precision Arithmetic & Gradient Descent'
output:
  html_document:
    df_print: paged
  html_notebook:
    code_folding: none
    highlight: textmate
---


```{r setup, include=FALSE}
source(file.path("..", "R", "colors.R"))
source(file.path("..", "R", "util.R"))

required_packages <- c('Matrix', 'testthat')
install_and_load_packages(required_packages)
```

# Exercise 0: Checking your understanding of floating point numbers and finite precision arithmetic

We observed in the class that the following code returns `FALSE`:

```{r, eval=FALSE}
a <- 1 + .Machine$double.eps
b <- 1 + .Machine$double.eps 
c <- 1
(a + b) + c == a + (b + c)
```

On the other hand, the following modification results in `TRUE`:

```{r, eval=FALSE}
b <- 0.5 + .Machine$double.eps
(a + b) + c == a + (b + c)
```

Why do we get the different result?

**Answer:**

Real numbers are encoded in 64-bit double-precision floating-point format in R. There is a 1-bit sign bit, an 11-bit exponent, and a 52-bit fraction. The real values of a given representation is $(-1)^{sign} (1+\sum_{i=1}^{52}b_{52-i}2^{-i}) \times 2^{e-1023}$, where $b_{i}$ is the $i^{th}$ bit in fraction and $e$ is the exponent. The finite-precision representation of real numbers can cause imprecision in arithmetic operations due to round-off errors.

In this example, `.Machine$double.eps` $=2^{-52}$ is the smallest difference between 1 and the next larger representable number. All the numbers here are positive, so I omit the sign bit in the following demonstration. The exponent and the fractional representation of all numbers are shown below.

In the first case, 

$$
\begin{align*}
a &= (1+2^{-52}) &\times 2^{1023-1023} \\
b &= (1+2^{-52}) &\times 2^{1023-1023} \\
c &= (1) &\times 2^{1023-1023} \\
a+b &= (1+2^{-52}) &\times 2^{1024-1023} \\
(a + b) + c &= (1+2^{-1}+2^{-52}) &\times 2^{1024-1023} \\
b + c  &\approx (1) &\times 2^{1024-1023} \\
a + (b + c) &\approx (1+2^{-1}) &\times 2^{1024-1023}
\end{align*}
$$

In the second case, 

$$
\begin{align*}
a &= (1+2^{-52}) &\times 2^{1023-1023} \\
b &= (1+2^{-51}) &\times 2^{1022-1023} \\
c &= (1) &\times 2^{1023-1023} \\
a+b &= (1+2^{-1}+2^{-51}) &\times 2^{1023-1023} \\
(a + b) + c &= (1+2^{-2}+2^{-52}) &\times 2^{1024-1023} \\
b + c &= (1+2^{-1}+2^{-52}) &\times 2^{1023-1023} \\
a + (b + c) &= (1+2^{-2}+2^{-52}) &\times 2^{1024-1023} \\
\end{align*}
$$
When the exponent is $k$, the minimum distinguishable difference is $2^{-52+k-1023}$. The minimum distinguishable difference in $b+c$ is $2^{-51}$ in the first case and is $2^{-52}$ in the second case. Thus, there is a round-off error on RHS in the first case but not the second. Since the calculation is accurate on LHS in both cases, this results in an evaluation of `FALSE` in the first case and a `TRUE` in the second case.


```{r, eval=FALSE, include=FALSE}
# adapted from https://stackoverflow.com/questions/50217954/double-precision-64-bit-representation-of-numeric-value-in-r-sign-exponent
Rcpp::cppFunction('void print_hex(double x) {
    uint64_t y;
    static_assert(sizeof x == sizeof y, "Size does not match!");
    std::memcpy(&y, &x, sizeof y);
    Rcpp::Rcout << std::hex << y << std::endl;
}', plugins = "cpp11", includes = "#include <cstdint>")
```


# Exercise 1: One-pass algorithm for calculating variance

Let's say you are running a Monte Carlo simulation to estimate the mean and variance of some random variable $\boldsymbol{\theta}$. 
Imagine that $\boldsymbol{\theta}$ is high-dimensional, so you don't want to keep all the Monte Carlo samples $\boldsymbol{\theta}_i$ due to the memory burden.
(If $\boldsymbol{\theta}$ is $10^6$-dimensional, for example, storing 1,000 Monte Carlo samples would require 8GB of memory.)
And it occurs to you that you can write a "one-pass" algorithm to calculate the sample mean and sample second moment of $\boldsymbol{\theta}$ by only keeping track of running totals of Monte Carlo samples. 
You can then, in theory, calculate the sample variance from these quantities.

The above idea is implemented and applied below. 
Try it and see what you get. 
Does the estimate make sense? If not, why?
How does the estimate change if you change the seed?

**Bonus question:** 
What is, roughly speaking, the distribution of the estimator `sample_var_with_question_mark` under repeated Monte Carlo simulations (i.e. without a fixed seed)?

```{r}
# original code
set.seed(140778)

running_sum <- 0
running_sum_sq <- 0

n_sample <- 10^3
for (i in 1:n_sample) {
  theta <- 1e10 + rnorm(1)
  running_sum <- running_sum + theta
  running_sum_sq <- running_sum_sq + theta^2
}
sample_mean <- running_sum / n_sample
sample_sec_mom <- running_sum_sq / n_sample

sample_var_with_question_mark <- (sample_sec_mom - (sample_mean)^2)
sample_var_with_question_mark
```

```{r}
# show the distribution of sample_var_with_question_mark

set.seed(10)
seed_rep_num = 1000
seeds = sample.int(1e8, seed_rep_num, replace=TRUE)

sample_var_with_question_mark_SAMPLE = rep(0, seed_rep_num)

for (curr_seed in 1:seed_rep_num) {
  set.seed(seeds[curr_seed])
  running_sum <- 0
  running_sum_sq <- 0
  
  n_sample <- 10^3
  for (i in 1:n_sample) {
    theta <- 1e10 + rnorm(1)
    running_sum <- running_sum + theta
    running_sum_sq <- running_sum_sq + theta^2
  }
  sample_mean <- running_sum / n_sample
  sample_sec_mom <- running_sum_sq / n_sample
  
  sample_var_with_question_mark <- (sample_sec_mom - (sample_mean)^2)
  sample_var_with_question_mark_SAMPLE[curr_seed] = sample_var_with_question_mark
}

plot(density(sample_var_with_question_mark_SAMPLE), 
     main="Empirical distribution of sample_var_with_question_mark")
```

```{r}
qqnorm(sample_var_with_question_mark_SAMPLE)
qqline(sample_var_with_question_mark_SAMPLE)
```


## Things to commit/submit

**Answer:**

I got `-147456` with the original seed $140778$. This does not make sense because $\theta \sim N(1e10, 1)$, so the sample variance should be approximately $1$.

The estimation is not accurate because arithmetic operations are not precise when real numbers are represented using the double-precision floating-point format. In `1e10 + rnorm(1)`, a relatively small number `rnorm(1)` is added to a very large number `1e10`. Since the minimum distinguishable difference decreases as the number becomes larger, this addition can cause round-off error and the precision in the `rnorm(1)` part is lost. In addition, the squaring of a large number in `theta^2` can cause additional loss in precision due to the same reason. As these round-off errors accumulate over 1000 iterations, the result becomes highly inaccurate. 

When the seed is changed, a difference sample is generated, so the sample variance changes. However, it is still far away from its theoretical value 1 because the round-off error remains.

The distribution of `sample_var_with_question_mark` without a fixed seed is roughly normal with mean 1 and a very large variance. We can think of the round-off errors as random noise added to the sample. Since both round-up and round-down errors exist, they approximately cancel each other. Moreover, the variance of the round-off error is large, as shown in the example. Thus, the round-off error is approximately normal with mean 0 and a large variance. Accordingly, `sample_var_with_question_mark` centers around 1 and has a very large variance.


# Exercise 2: Generating random real numbers in finite precision arithmetic

## Part A

You learned in the class that numeric values in computers are typically represented with 64-bits double-precision numbers, having 52-bits fraction. 
In particular, a "Uniform(1, 2)" random variable in double-precision generates a number from the $2^{52}$ unique values between $1$ and $2$.
The amount of randomness in it, therefore, is equivalent to 52 Bernoulli random variables.
Write a function to generate a Uniform(1, 2) random variable in double-precision with randomness coming only from the `rbinom()` function.
Check the function's behavior by plotting a histogram of random variables generated from it. 

```{r, eval=TRUE}
double_prec_runif <- function () {
  # generate a Uniform(1, 2) random variable
  fraction_part = rbinom(n=52, size=1, prob=0.5)
  random_number = 1
  for (i in 1:52) {
    if (fraction_part[i]) {
      random_number = random_number + 2^(-i)
    }
  }
  return(random_number)
}

n_draw <- 10^5
unif_rv <- sapply(1:n_draw, function (i) double_prec_runif())

rv_hist <- hist(unif_rv, breaks = 21, plot = FALSE)
plot(
  rv_hist,
  freq = FALSE,
  xlab = "Realized values",
  col = "#bdcedb", # JHU Spirit Blue with saturation and luminance adjusted: https://hslpicker.com/#bdcedb
  border = "#859bad" # Another variation of JHU Spirit Blue
)
```

## Part B

As we observed above, a Uniform(1, 2) random variable in double-precision takes one of the $2^{52}$ unique values.
Two independently drawn random variables, therefore, has small but positive probability of having exactly the same value.
We can calculate the expected number of duplicate values in i.i.d. Uniform(1, 2) random variables as follows.
When drawing with replacement $d$ independent random variables out of $n$ choices, the probability of the $i$-th element never selected is $(1 - n^{-1})^d$ for $i = 1, \ldots, n$. 
The expected number of unique items drawn, therefore, is $n (1 - n^{-1})^d$.
In turn, the expected number of duplicates is $d - n (1 - n^{-1})^d$.
For the case of double-precision uniform random variables, we have $n = 2^{52}$ for the number of choices.

The `calc_num_expected_duplicate()` function below implements the above formula for the expected number of duplicates. 
When you call `calc_num_expected_duplicate(n_draw = 10^6, n_choice = 2^52)` to calculate the expected number for $10^6$ Uniform(1, 2) random variables, however, the function returns _exact zero_ due to a finite precision error. 
What is the issue?

```{r}
calc_num_expected_duplicate <- function (n_draw, n_choice) {
  # To be remedied.
  prob_item_never_chosen <- (1 - 1 / n_choice) ^ n_draw
  print(prob_item_never_chosen)
  print(prob_item_never_chosen == 1)
  n_expected_unique <- n_choice * (1 - prob_item_never_chosen)
  return(n_draw - n_expected_unique)
}
calc_num_expected_duplicate(n_draw=1e6, n_choice=2^(52)) == 0
```

**Answer:**

  1. In this example, `1 - 1 / n_choice` can be represented accurately with the double-precision floating-point format. However, it is very close to 1, so taking an exponential of it is still close to 1. This results in an round-off error because the exponent part of the number representation does not change much, and so does the minimum distinguishable difference. Any arithmetic changes smaller than the minimum distinguishable difference is rounded-off. Since `n_draw` is large, this leads to a loss of precision. 

  2. Moreover, in `n_choice * (1 - prob_item_never_chosen)`, `n_choice` is much larger than `1 - prob_item_never_chosen`, the latter of which is close to $0$. Multiplying a small number with a large number causes the minimum distinguishable difference to decrease. This round-off error further contribute to the loss of precision.


Formulas involving $1 + x$ for $|x| \ll 1$ ($x = - n^{-1}$ in the above case) occur commonly enough in statistics/scientific computing that R provides a function `log1p(x)` for computing $\log(1 + x)$ to deal with one such situation.
In more general calculations involving $|x| \ll 1$ (e.g. for a "sinc" funciton $\operatorname{sinc}(x) = \sin(x) / x$), a Taylor expansion provides a general approach to control truncation errors.

Remedy the `calc_num_expected_duplicate()` function by applying the second-order Taylor expansion to $(1 + x)^d$ and simplifying the formula $d - x^{-1} (1 - x)^d$ to avoid truncation errors. 
(Remark: i) in this specific case, you could use the binomial expansion and calculate it exactly; 
ii) applying the second-order expansion to $(1 + x)^d$ corresponds to applying the first-order expansion to the overall expression $d - x^{-1} (1 - x)^d$.)
Use the more numerically accurate version of `calc_num_expected_duplicate()` to calculate the expected number of duplicates among $10^6$ Uniform(1, 2) random variables.

**Answer:**

Let $f(x) = x^d$. Then the second-order Taylor expansion of $f(\cdot)$ around $x=1$ is 

$$
\begin{align*}
f(1+x) &\approx f(1) + f'(1)x + \frac{x^2}{2}f''(1) \\
       &= 1 + dx + \frac{x^2}{2}d(d-1)
\end{align*}
$$

The original formula is simplified as

$$
\begin{align*}
d - n \left(1-(1-\frac{1}{n})^d\right) &=
d - n + n(1-\frac{1}{n})^d \\
&\approx d - n + n(1 - \frac{d}{n} + \frac{1}{2n^2}d(d-1)) \\
&= d - n + n - d + \frac{1}{2n}d(d-1) \\
&= \frac{1}{2n}d(d-1)
\end{align*}
$$

We can use log transformation to increase the accuracy in computation.

$$d - n \left(1-(1-\frac{1}{n})^d\right) = exp\left(log(d) + log(d-1) - log(n) - log(2)\right)$$

```{r}
# Remedied using second-order Taylor expansion
calc_num_expected_duplicate <- function (n_draw, n_choice) {
  result = log(n_draw) + log(n_draw - 1) - log(n_choice) - log(2)
  return(exp(result))
}
calc_num_expected_duplicate(n_draw=1e6, n_choice=2^(52))
```

## Part C

The expected number from Part B, if you calculated it correctly, suggests that it is extremely rare to get duplicate values when generating $10^6$ uniform random variables in double-precision.
This is reassuring because we routinely generate a far larger number of random numbers in modern applications.
If we use R's `runif()` function, however, this is _not_ what we observe because [R's random numbers are generated using 32-bit integers](https://www.tandfonline.com/doi/full/10.1080/00031305.2020.1782261).
Check the number of duplicates from `runif(10^6)` against the expected value for a floating-point number with 32-bits of precision. 
Check also that, using your implementation of random uniform from Part A, duplicates are indeed rare in double-precision.

```{r, eval=TRUE}
n_draw <- 10^6
base_r_unif <- runif(n_draw, min = 1, max = 2)
unif_rv <- sapply(1:n_draw, function (i) double_prec_runif())

# Check the number of duplicates against the expected value
cat(sprintf("The number of duplicates from runif(10^6) is %d\n", n_draw - length(unique(base_r_unif))))
cat(sprintf("The expected number of duplicates from runif(10^6) is %.4f\n", calc_num_expected_duplicate(n_draw=n_draw, n_choice=2^(32))))
cat(sprintf("The number of duplicates from double_prec_runif() is %d\n", n_draw - length(unique(unif_rv))))
cat(sprintf("The expected number of duplicates from runif(10^6) is %.4f\n", calc_num_expected_duplicate(n_draw=n_draw, n_choice=2^(52))))
```


## Things to commit/submit
Answers to the tasks/questions above.


# Exercise 3: Numerical differentiation

## Part A: Potential caveats of numerical differentiation

You have learned in the class that the centered difference $\frac{f(x + \Delta x) - f(x - \Delta x)}{2 \Delta x}$ approximates the derivative $f'(x)$ up to an $O(\Delta x^2)$ error.
To numerically verify this, we design the following experiment using a log of logistic function (i.e. $f(x) = \log\{ \exp(x) / (1 + \exp(x)) \}$) as an example.
We evaluate its derivative using an analytical formula and compare it to the centered difference approximations as $\Delta x$ varies from $2^{-1}$ to $2^{-52}$.
Using the starter code provided below, plot the relative error of the numerical approximation in a log-log scale.
What do you expect the slope of the plotted line to be?

**Answer:** I expect the plot on the log-log scale to be a straight line that has a slope of -2. Moreover, as $\Delta x$ tends to 0 and $log \Delta x$ tends to $-\infty$, the error goes to 0 and the log error goes to $-\infty$.

The centered difference approximation is $\frac{f(x + \Delta x) - f(x - \Delta x)}{2 \Delta x} \approx f'(x) + \frac{(\Delta x)^2f'''(x)}{6} = f'(x) + \Theta((\Delta x)^2)$. Thus, the error is $\epsilon \approx \Theta((\Delta x)^2)$.

By definition of big-theta notation, there exists constants $c_1$, $c_2$, $n_0 > 0$ such that for all $n > n_0$, $c_1(\Delta x)^2 \leq \epsilon \leq c_2(\Delta x)^2$. This indicates that $c'_1 + 2log_2(\Delta x) \leq log_2 \epsilon \leq c'_2 + 2log_2(\Delta x)$. For $\Delta x = 2^{-k}$ and $k=1,\dots, 52$, $c'_1 - 2k \leq log_2 \epsilon \leq c'_2 - 2k$. Thus, the slope on the log-log scale is approximately -2.

```{r, eval=TRUE}
#' Approximate derivative via centered-difference approximation
approximate_deriv <- function(func, x, dx) {
  # (f(x+dx) - f(x-dx)) / (2*dx)
  return((func(x+dx) - func(x-dx)) / (2*dx))
}

log_logistic_func <- function(x) {
  # f(x)=log{exp(x)/(1+exp(x))} = x - log(1+exp(x))
  return(x - log(1+exp(x)))
}

log_logistic_deriv <- function(x) {
  # f'(x) = 1 / (1+exp(x))
  return(1 / (1+exp(x)))
}

# Calculate and plot the relative errors of finite difference approximations
set.seed(615)
x <- rnorm(1)

log2_dx <- - seq(1, 52)
numerical_deriv <- sapply(
  2^log2_dx, function(dx) approximate_deriv(log_logistic_func, x, dx)
)
analytical_deriv <- log_logistic_deriv(x)
rel_err <- abs(
  (analytical_deriv - numerical_deriv) / analytical_deriv
)

fontsize <- 1.2
plot(
  log2_dx, log2(rel_err),
  frame = FALSE, # Remove the ugly box
  col = jhu_color$spiritBlue,
  xlab = "log2 stepsize for finite difference",
  ylab = "log2 rel errors",
  xlim = rev(range(log2_dx)),
  cex.lab = fontsize,
  cex.axis = fontsize,
  type = 'l'
)

```

**Questions:**
Does the relative error plot agree with what the mathematical theory predicts? Explain the observed phenomenon.

Also, what is the slope in the "latter part" of your plotted line, roughly speaking? 
Explain why we observe such a slope in the relative error. 
(Hint: Consider a machine with 10 decimal digits of accuracy. What happens if we numerically differentiate $f(x) = x$ at $x = 1$ with $\Delta x = 0.1, 0.01, 0.001, \ldots$.)

**Answer:** The relative error plot partially agrees with the theory. It is approximately a straight line with slope -2 when $\Delta x$ is larger than $2^{-16}$. However, as $\Delta x$ further decreases, log error increases linearly with a slope 1 and some fluctuations.

When $\Delta x$ is small, $f(x+\Delta x) - f(x-\Delta x))$ is close to 0. This small difference cannot be accurately represented because real numbers are stored using the double-precision floating-point format. The lose of precision is much larger than the approximation error in the centered difference. As a result, even though the theoretical approximation error decreases for smaller $\Delta x$'s, the lose of precision increases and contributes to an increase of the total error. This leads to the increasing pattern when $\Delta x$ is very small.

Suppose $\Delta x = 2^{-k}$. The centered difference is $\frac{f(x + \Delta x) - f(x - \Delta x)}{2 \Delta x} = \frac{diff_k}{2 \times 2^{-k}} = \frac{diff_k/2}{2 \times 2^{-(k+1)}} \approx \frac{diff_{k+1}}{2 \times 2^{-(k+1)}}$. Thus, when $\Delta x$ decreases from $2^{-k}$ to $2^{-(k+1)}$, the difference $f(x + \Delta x) - f(x - \Delta x)$ should also decrease by approximately half. For a difference close to zero, halving causes the lose of the last significant digit in its binary expansion when calculating $f(x + \Delta x)$ and $f(x - \Delta x)$. This causes the relative error to double. Thus, the relative error $\epsilon(k)$ as a function of $k$ is approximately $\epsilon(k) \propto 2^k \Rightarrow log_2 \epsilon(k) \approx \text{const} + k$, which is consistent with the slope 1 in the latter part of the plot.

In the latter part of the plot, the log error increases with fluctuations because the amount of round-off error depends on the last few digits for $f(x + \Delta x) - f(x - \Delta x)$ in its binary expansion. Though the round-off error is expected to be doubled as $k$ increases by 1, the actual error may increase, remain the same, or even have a slight decrease depending on whether a round-up or a round-down error occurs. Thus, the log error increases with fluctuations.

## Part B: Extention to multivariate functions

As mentioned in the lecture, the finite difference method extends to a function with multivariate input $f: \mathbb{R}^d \to \mathbb{R}$ and can be used to approximate the gradient $\nabla f(\boldsymbol{x})$.
The extension relies on the fact $f(\boldsymbol{x} + \Delta x \, \boldsymbol{e}_i) = f(\boldsymbol{x}) + \Delta x \, \partial_i f(\boldsymbol{x}) + \ldots$ for $i = 1, \ldots, d$.

Using the starter code provided below, implement a function to approximate $\nabla f(\boldsymbol{x})$ via the centered difference method.
To test this function, take as an example a quadratic function $f(\boldsymbol{x}) = - \boldsymbol{x}^\intercal \boldsymbol{\Phi} \boldsymbol{x} / 2$ with positive definite $\boldsymbol{\Phi}$, which can be thought of as a log-likelihood function for $\boldsymbol{x} \sim \mathscr{N}(\boldsymbol{0}, \boldsymbol{\Phi})$ up to an additive constant.
First show that its gradient is given by $\nabla f(\boldsymbol{x}) = - \boldsymbol{\Phi} \boldsymbol{x}$; 
more generally, the gradient of a quadratic function $g(\boldsymbol{x}) = \boldsymbol{x}^\intercal \boldsymbol{A} \boldsymbol{x}$ is given by $\nabla g(\boldsymbol{x} = (\boldsymbol{A} + \boldsymbol{A}^\intercal) \boldsymbol{x}$ for any square matrix $\boldsymbol{A}$.
Then compare a numerically approximated gradient to one computed via an analytical formula.

**Answer:**

Suppose $\boldsymbol{x} \in \mathbb{R}^n$ and $\boldsymbol{\Phi} \in \mathbb{R}^{n\times n}$. Note that the positive definite matrix $\boldsymbol{\Phi}$ is symmetric.

The analytical form of $\nabla f(\boldsymbol{x})$ is as follows:

$$
\begin{align*}
f(\boldsymbol{x}) &= - \boldsymbol{x}^\intercal \boldsymbol{\Phi} \boldsymbol{x} / 2 = -\frac{1}{2} \sum_{i=1}^n \sum_{j=1}^n \boldsymbol{x}_i \boldsymbol{x}_j \boldsymbol{\Phi}_{ij} \\
\frac{\partial f(\boldsymbol{x})}{\partial \boldsymbol{x}_k} &= -\left(\sum_{i \in {1, \dots, n}, i \neq k} \boldsymbol{x}_i \boldsymbol{\Phi}_{ik}\right) -\boldsymbol{x}_k \boldsymbol{\Phi}_{kk} = -\sum_{i=1}^n \boldsymbol{x}_i \boldsymbol{\Phi}_{ik}\\
\nabla f(\boldsymbol{x}) &= -\boldsymbol{\Phi}\boldsymbol{x}
\end{align*}
$$

The approximation of $\nabla f(\boldsymbol{x})$ through centered difference method is as follows:

$$
\begin{align*}
f(\boldsymbol{x} + \Delta x \boldsymbol{e}_k) 
&\approx f(\boldsymbol{x}) + \sum_{i=1}^n \frac{\partial f(\boldsymbol{x})}{\partial \boldsymbol{x}_i} (\Delta x \boldsymbol{e}_k)_i  \\
&\quad + \frac{1}{2} \sum_{i=1}^n  \sum_{j=1}^n\frac{\partial^2 f(\boldsymbol{x})}{\partial \boldsymbol{x}_i \partial \boldsymbol{x}_j} (\Delta x \boldsymbol{e}_k)_i (\Delta x \boldsymbol{e}_k)_j \\
&\quad + \frac{1}{6} \sum_{i=1}^n  \sum_{j=1}^n \sum_{k=1}^n \frac{\partial^3 f(\boldsymbol{x})}{\partial \boldsymbol{x}_i \partial \boldsymbol{x}_j \partial \boldsymbol{x}_k} (\Delta x \boldsymbol{e}_k)_i (\Delta x \boldsymbol{e}_k)_j (\Delta x \boldsymbol{e}_k)_k \\
&= f(\boldsymbol{x}) + \frac{\partial f(\boldsymbol{x})}{\partial \boldsymbol{x}_k} \Delta x 
+ \frac{1}{2} \frac{\partial^2 f(\boldsymbol{x})}{\partial \boldsymbol{x}_k^2} (\Delta x)^2 
+ \frac{1}{6} \frac{\partial^3 f(\boldsymbol{x})}{\partial \boldsymbol{x}_k^3} (\Delta x)^3
\\
\frac{\partial f(\boldsymbol{x})}{\partial \boldsymbol{x}_k} &\approx \frac{f(\boldsymbol{x}_k + \Delta x \boldsymbol{e}_k) - f(\boldsymbol{x}_k - \Delta x \boldsymbol{e}_k)}{2 \Delta x} + \Theta((\Delta x)^2)
\end{align*}
$$

More generally, for $g(\boldsymbol{x}) = \boldsymbol{x}^\intercal \boldsymbol{A} \boldsymbol{x}$,

$$
\begin{align*}
g(\boldsymbol{x}) &= \boldsymbol{x}^\intercal \boldsymbol{A} \boldsymbol{x}
= \sum_{i=1}^n \sum_{j=1}^n \boldsymbol{x}_i \boldsymbol{x}_j \boldsymbol{A}_{ij} \\
\frac{\partial g(\boldsymbol{x})}{\partial \boldsymbol{x}_k} 
&= \left(\sum_{i \in {1, \dots, n}, i \neq k} \boldsymbol{x}_i \boldsymbol{A}_{ik}\right)
+ \left(\sum_{i \in {1, \dots, n}, i \neq k} \boldsymbol{x}_i \boldsymbol{A}_{ki}\right) 
+2\boldsymbol{x}_k \boldsymbol{A}_{kk} 
= \sum_{i=1}^n \boldsymbol{x}_i (\boldsymbol{A}_{ki} + \boldsymbol{A}_{ik}) \\
\nabla g(\boldsymbol{x}) &= (\boldsymbol{A} + \boldsymbol{A}^\intercal) \boldsymbol{x} \\
\frac{\partial g(\boldsymbol{x})}{\partial \boldsymbol{x}_k}
&\approx  \frac{g(\boldsymbol{x}_k + \Delta x \boldsymbol{e}_k) - g(\boldsymbol{x}_k - \Delta x \boldsymbol{e}_k)}{2 \Delta x} + \Theta((\Delta x)^2)
\end{align*}
$$


```{r, eval=TRUE}
approx_grad <- function(func, x, dx = .Machine$double.eps^(1/3)) {
  numerical_grad <- rep(0, length(x))
  for (i in 1:length(x)) {
    plus_dx = x
    plus_dx[i] = plus_dx[i] + dx
    minus_dx = x
    minus_dx[i] = minus_dx[i] - dx
    numerical_grad[i] = (func(plus_dx) - func(minus_dx)) / (2 * dx)
  }
  return(numerical_grad)
}

set.seed(410)
n_param <- 4
X <- matrix(rnorm(2 * n_param^2), nrow = 2 * n_param, ncol = n_param)
Sigma_inv <- t(X) %*% X
  
#' Calculate log-density of centered Gaussian up to an additive factor
gaussian_logp <- function(x, Sigma_inv) {
  logp <- - .5 * t(x) %*% Sigma_inv %*% x
  return(logp)
}

gaussian_grad <- function(x, Sigma_inv) {
  grad = -Sigma_inv %*% matrix(x, ncol=1)
  grad = as.vector(grad)
  return(grad)
}

x <- c(3, 1, 4, 1)

analytical_grad <- gaussian_grad(x, Sigma_inv)
numerical_grad <- approx_grad(function(x) gaussian_logp(x, Sigma_inv), x)
testthat::expect_true(are_all_close(
  analytical_grad, numerical_grad, abs_tol = Inf, rel_tol = 1e-3
))
```


## Things to commit/submit
For Part A, completed code and answers to the questions.
For Part B, derivation of the gradient and completed code that passes the test.


# Exercise 4: Continuing development of `hiperglm` &mdash; finding MLE via `stats::optim()` and testing it

We will continue the development of the `hiperglm` package. 

As discussed, our plan is to first implement an MLE finder via a pre-packaged general-purpose BFGS optimizer, against which we can then use to benchmark our custom implementation of iteratively reweighted least squares. 
The idea is that using the pre-packaged function is less error-prone, especially since the gradient function is easy to unit-test.

To be extra safe, we are going to test the MLE finder via BFGS against an even simpler algorithm: the least square for a linear model via the analytical expression $\hat{\boldsymbol{\beta}}_\textrm{mle} = (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \boldsymbol{y}$.
Complete the following steps toward this goal.
Remember that you can use [`devtools::load_all()`](https://r-pkgs.org/code.html#sec-code-load-all) to help your [package development process](https://r-pkgs.org/code.html#constant-health-checks). 

0. Make and check out (`git branch` and `git checkout`) a development branch (call it `dev` or `develop`) where you will commit the work below.
1. Set up the use of `testthat`.
2. Copy the helper functions `are_all_close()` and `simulate_data()` from the instructor's repo.
3. Test the `are_all_close()` function. 
Write a test to cover at least three cases: the function 1) correctly returns `TRUE`, 2) correctly returns `FALSE` because the relative error is above `rel_tol`, and 3) correctly returns `FALSE` because the absolute error is above `abs_tol`. 
Make sure that the logic of the test is clear and easy to understand &mdash; tests are there to clarify and validate the code's behavior; unreadable tests defeat the purpose.
<br> _Remark:_ This is just an exercise, but in general there is nothing strange about testing a function you use to test another function. It is perfectly reasonable to consider testing any functions complicated enough to potentially contain insiduous bugs.
4. Set up the (currently failing) test to compare the MLE estimated via pseudo-inverse and via BFGS. 
5. Implement the MLE finder via peudo-inverse. 
Please, pretty please, do not calculate an inverse of a matrix. 
(Otherwise, it's fine to use `solve()` instead of `chol()`.)
6. Implement the MLE finder via BFGS:
    a. Implement functions to calculate the log-likelihood and gradient under a linear model. 
    This function technically requires an estimate of the residual variance, but its value does not affect the MLE for the coefficients and hence is irrelevant for our purpose. 
    You can just specify it as an optional parameter e.g. as `function(..., noise_var = 1)`.
    b. Test your gradient calculation by comparing it against a numerical one.
    For this purpose, use the function from Exercise 2, Part B to approximate the gradient of a given function via finite difference.
    <br> _Remark:_ Technically, this test checks only for consistency between the log-likelihood and gradient function.
    It is possible to have both functions wrong but consistent with each other so that the test passes.
    This test nonetheless can catch many bugs and, since the gradient calculation tends to be more error prone, it's good to make sure there is no mistake there. 
    c. Plug in the now tested-and-true log-likelihood and gradient functions to `stats::optim()` to find MLE.
7. Run `devtools::test()` and check that the then-failing test from Step 2 now passes.
8. Open a pull request from your development branch to the main/master one. 
    
## Things to commit/submit

Link to the pull request. <https://github.com/Wenxuan-Lu/hiperglm/pull/1>
