---
title: 'Homework: Finite Precision Arithmetic & Gradient Descent'
output:
  html_document:
    df_print: paged
  html_notebook:
    code_folding: none
    highlight: textmate
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
source(file.path("..", "R", "colors.R"))
source(file.path("..", "R", "util.R"))

required_packages <- c('Matrix', 'testthat')
install_and_load_packages(required_packages)
```

# Exercise 0: Checking your understanding of floating point numbers and finite precision arithmetic

We observed in the class that the following code returns `FALSE`:

```{r, eval=FALSE}
a <- 1 + .Machine$double.eps
b <- 1 + .Machine$double.eps 
c <- 1
(a + b) + c == a + (b + c)
```

On the other hand, the following modification results in `TRUE`:

```{r, eval=FALSE}
b <- 0.5 + .Machine$double.eps
(a + b) + c == a + (b + c)
```

Why do we get the different result?

**Explanation**: An important thing about double precision is that there are $2^{52}$ machine-representable values between two reals $2^k$ and $2^{k+1}$, with increment $2^k \times 2^{52}$. This means $2=2^1$ is not recognized as different from $2^1 + 2^{-52}$, but is different from $2^1 + 2^1 \times 2^{-52}$ in machine perspective. Shown as follows.

Let $e = 2^{-52}$.

```{r}
2 + .Machine$double.eps == 2 # TRUE, 2 = 2 + 2^-52
2 + .Machine$double.eps * 2 == 2 # FALSE, 2 != 2 + 2*2^-52 = 2 + 2^-51
```

Similarly, $1 = 2^0$ is recognized as different from $2^0 + 2^0 \times 2^{-52}$, but not different $1 + 2^{-1} \times 2^{-52}$

```{r}
1 + .Machine$double.eps == 1  # FALSE
1 + .Machine$double.eps / 2 == 1 # TRUE
```


Case 1: in the first example example evaluated to false

- on the LHS, $a + b = 1 + e + 1 + e$ is evaluated to $2 + 2e$, because $2$ is different from $2+2e$. $(a + b) + c = 2 + 2e + 1$ is thus evaluated to $3 + 2e$.
- While on the RHS, $b + c = 1 + e + 1 = 2 + e$ evaluated to $2$, and $a + (b + c)$ is thus evaluated as $1 + e + 2 = 3 + e$ which is evaluated as $3$.
- Thus LHS $\neq$ RHS

Case 2: However, in the case of $b=0.5$, we have equality

- $a = 1+e \neq 1$ as proven above 
- $b = 0.5 + e \neq 0.5$ since the lowest it can be separated from is $2^{-1} + 2^{-1} \times e$ (shown below). Therefore, 
- on the LHS, $a+b = 1 + e + 0.5 + e = 1.5  +2e$, $(a+b) + c = 1.5 + 2e + 1 = 2.5 + 2e$. 
- on the RHS, $a + (b+c) = 1 + e + (0.5 + e + 1) = 1 + e + 1.5 + e = 2.5 + 2e$. 
- Therefore, LHS $==$ RHS

```{r}
0.5 + .Machine$double.eps == 0.5 # FALSE, 2^-1 + 2^0 * e != 2^-1 
0.5 + .Machine$double.eps / 2 == 0.5 # FALSE, 2^-1 + 2^-1 * e != 2^-1 
0.5 + .Machine$double.eps / 4 == 0.5 # TRUE, 2^-1 + 2^-2 * e == 2^-1 
```


# Exercise 1: One-pass algorithm for calculating variance

Let's say you are running a Monte Carlo simulation to estimate the mean and variance of some random variable $\boldsymbol{\theta}$.
Imagine that $\boldsymbol{\theta}$ is high-dimensional, so you don't want to keep all the Monte Carlo samples $\boldsymbol{\theta}_i$ due to the memory burden.
(If $\boldsymbol{\theta}$ is $10^6$-dimensional, for example, storing 1,000 Monte Carlo samples would require 8GB of memory.) And it occurs to you that you can write a "one-pass" algorithm to calculate the sample mean and sample second moment of $\boldsymbol{\theta}$ by only keeping track of running totals of Monte Carlo samples.
You can then, in theory, calculate the sample variance from these quantities.

The above idea is implemented and applied below.
Try it and see what you get.
Does the estimate make sense?
If not, why?
How does the estimate change if you change the seed?

**Bonus question:** What is, roughly speaking, the distribution of the estimator `sample_var_with_question_mark` under repeated Monte Carlo simulations (i.e. without a fixed seed)?



```{r}
set.seed(140778)

running_sum <- 0
running_sum_sq <- 0

n_sample <- 10^3
for (i in 1:n_sample) {
  theta <- 1e10 + rnorm(1)
  running_sum <- running_sum + theta
  running_sum_sq <- running_sum_sq + theta^2
}
sample_mean <- running_sum / n_sample
sample_sec_mom <- running_sum_sq / n_sample

sample_var_with_question_mark <- (sample_sec_mom - (sample_mean)^2)
```

## Things to commit/submit

Answers to the questions above: 

No, the estimate does not make sense. In the estimation, the code goes `theta <- 1e10 + rnorm(1)`, which means $\theta \sim N(10^{-10}, 1)$. The expected variance is 1, but here we get the estimated sample variance to be `-147456`.

```{r}
sample_var_with_question_mark
```

*Explanation*: `sample_var_with_question_mark` is trying to estimate this variance of $\theta=1$. However, due to finite precision, this estimate can vary a lot around $1$. `1e10 + rnorm(1)` is a large number plus a small error term $\epsilon$, and when we are squaring it in calculating the second moment, this large number $+ \epsilon$ squared requires more than 52 fraction bits to represent. Therefore, machine cannot represent this number accurately. $\theta^2$ gets rounded up or rounded down by the machine; the estimated sample variance is no longer accurate at $1$, but can vary around 1 by a lot. 

If we change the seed, the estimated variance `sample_var_with_question_mark` can vary a lot around 1, both positive and negative.

The distribution of the estimator `sample_var_with_question_mark` under repeated Monte Carlo simulations is normal distribution with mean 1 by central limit theorem. Below we obtain 1000 sample variance values from 1000 simulation, and get a histogram that resembles normally distribution with large variance.

```{r}
test <- function(i) {
  running_sum <- 0
  running_sum_sq <- 0

  n_sample <- 10^3
  for (i in 1:n_sample) {
    theta <- 1e10 + rnorm(1)
    running_sum <- running_sum + theta
    running_sum_sq <- running_sum_sq + theta^2
  }
  sample_mean <- running_sum / n_sample
  sample_sec_mom <- running_sum_sq / n_sample

  sample_var_with_question_mark <- (sample_sec_mom - (sample_mean)^2)
  sample_var_with_question_mark
}

n_draw <- 1000
var_rv <- sapply(1:n_draw, function (i) test())

hist(var_rv)
```



# Exercise 2: Generating random real numbers in finite precision arithmetic

## Part A

You learned in the class that numeric values in computers are typically represented with 64-bits double-precision numbers, having 52-bits fraction.
In particular, a "Uniform(1, 2)" random variable in double-precision generates a number from the $2^{52}$ unique values between $1$ and $2$.
The amount of randomness in it, therefore, is equivalent to 52 Bernoulli random variables.
Write a function to generate a Uniform(1, 2) random variable in double-precision with randomness coming only from the `rbinom()` function.
Check the function's behavior by plotting a histogram of random variables generated from it.

```{r}
double_prec_runif <- function () {
  # Fill in
  1 + sum(rbinom(52, 1, 0.5) * 2^(- seq(1, 52)))
}

n_draw <- 10^5
unif_rv <- sapply(1:n_draw, function (i) double_prec_runif())

rv_hist <- hist(unif_rv, breaks = 21, plot = FALSE)
plot(
  rv_hist,
  freq = FALSE,
  xlab = "Realized values",
  col = "#bdcedb", # JHU Spirit Blue with saturation and luminance adjusted: https://hslpicker.com/#bdcedb
  border = "#859bad" # Another variation of JHU Spirit Blue
)
```

## Part B

As we observed above, a Uniform(1, 2) random variable in double-precision takes one of the $2^{52}$ unique values.
Two independently drawn random variables, therefore, has small but positive probability of having exactly the same value.
We can calculate the expected number of duplicate values in i.i.d. Uniform(1, 2) random variables as follows.
When drawing with replacement $d$ independent random variables out of $n$ choices, the probability of the $i$-th element never selected is $(1 - n^{-1})^d$ for $i = 1, \ldots, n$.
The expected number of unique items drawn, therefore, is $n [1 - (1 - n^{-1})^d]$.
In turn, the expected number of duplicates is $d - n [1 - (1 - n^{-1})^d]$.
For the case of double-precision uniform random variables, we have $n = 2^{52}$ for the number of choices.

The `calc_num_expected_duplicate()` function below implements the above formula for the expected number of duplicates.
When you call `calc_num_expected_duplicate(n_draw = 10^6, n_choice = 2^52)` to calculate the expected number for $10^6$ Uniform(1, 2) random variables, however, the function returns *exact zero* due to a finite precision error.
What is the issue?

```{r}
calc_num_expected_duplicate <- function (n_draw, n_choice) {
  # To be remedied.
  prob_item_never_chosen <- (1 - 1 / n_choice) ^ n_draw
  n_expected_unique <- n_choice * (1 - prob_item_never_chosen)
  return(n_draw - n_expected_unique)
}
calc_num_expected_duplicate(n_draw = 10^6, n_choice = 2^52)
```


**Answer**: This is underflow due to finite precision. The formula $d - n [1 - (1 - n^{-1})^d]$ involves calculating $2^{-52}$ to the power of $10^6$, which is smaller than `.Machine$double.xmin` (see below). So the best the machine can represent the answer is $0$

```{r}
.Machine$double.xmin
(2^(-52))^(10^6)
```

Formulas involving $1 + x$ for $|x| \ll 1$ ($x = - n^{-1}$ in the above case) occur commonly enough in statistics/scientific computing that R provides a function `log1p(x)` for computing $\log(1 + x)$ to deal with one such situation.
In more general calculations involving $|x| \ll 1$ (e.g. for a "sinc" funciton $\operatorname{sinc}(x) = \sin(x) / x$), a Taylor expansion provides a general approach to control truncation errors.

Remedy the `calc_num_expected_duplicate()` function by applying the second-order Taylor expansion to $(1 - x)^d$ and simplifying the formula $d - x^{-1} [1 - (1 - x)^d]$ to avoid truncation errors.
(Remark: i) in this specific case, you could use the binomial expansion and calculate it exactly; ii) applying the second-order expansion to $(1 - x)^d$ corresponds to applying the first-order expansion to the overall expression $d - x^{-1} [1 - (1 - x)^d]$.) Use the more numerically accurate version of `calc_num_expected_duplicate()` to calculate the expected number of duplicates among $10^6$ Uniform(1, 2) random variables.

```{r}
calc_num_expected_duplicate <- function (n_draw, n_choice) {
  # remedied.
  x <- 1/n_choice
  0.5 * n_draw * (n_draw - 1) * x
}
calc_num_expected_duplicate(n_draw = 10^6, n_choice = 2^52)
```



## Part C

The expected number from Part B, if you calculated it correctly, suggests that it is extremely rare to get duplicate values when generating $10^6$ uniform random variables in double-precision.
This is reassuring because we routinely generate a far larger number of random numbers in modern applications.
If we use R's `runif()` function, however, this is *not* what we observe because [R's random numbers are generated using 32-bit integers](https://www.tandfonline.com/doi/full/10.1080/00031305.2020.1782261).
Check the number of duplicates from `runif(10^6)` against the expected value for a floating-point number with 32-bits of precision.
Check also that, using your implementation of random uniform from Part A, duplicates are indeed rare in double-precision.


```{r}
n_draw <- 10^6
base_r_unif <- runif(n_draw, min = 1, max = 2)
unif_rv <- sapply(1:n_draw, function (i) double_prec_runif())
# Check the number of duplicates against the expected value
calc_num_expected_duplicate(n_draw = 10^6, n_choice = 2^32)

sum(base_r_unif %in% unif_rv) # duplicates in my implementation in part A
```

The number of duplicates from `runif(10^6)` against the expected value for a floating-point number with 32-bits of precision is **116.4152**.

Using my implementation of random uniform from Part A, the number of duplicates is 0. Duplicates are rare in double precision


## Things to commit/submit

Answers to the tasks/questions above.

# Exercise 3: Numerical differentiation

## Part A: Potential caveats of numerical differentiation

You have learned in the class that the centered difference $\frac{f(x + \Delta x) - f(x - \Delta x)}{2 \Delta x}$ approximates the derivative $f'(x)$ up to an $O(\Delta x^2)$ error.
To numerically verify this, we design the following experiment using a log of logistic function (i.e. $f(x) = \log\{ \exp(x) / (1 + \exp(x)) \}$) as an example.
We evaluate its derivative using an analytical formula and compare it to the centered difference approximations as $\Delta x$ varies from $2^{-1}$ to $2^{-52}$.
Using the starter code provided below, plot the relative error of the numerical approximation in a log-log scale.
What do you expect the slope of the plotted line to be?

```{r}
#' Approximate derivative via centered-difference approximation
approximate_deriv <- function(func, x, dx) {
  # Fill in
  (func(x + dx) - func(x - dx)) / (2*dx)
}

log_logistic_func <- function(x) {
  # Fill in
  log(exp(x) / (1 + exp(x)))
}

log_logistic_deriv <- function(x) {
  # Fill in
  1/ (1 + exp(x))
}

# Calculate and plot the relative errors of finite difference approximations
set.seed(615)
x <- rnorm(1)

log2_dx <- - seq(1, 52)
numerical_deriv <- sapply(
  2^log2_dx, function(dx) approximate_deriv(log_logistic_func, x, dx)
)
analytical_deriv <- log_logistic_deriv(x)
rel_err <- abs(
  (analytical_deriv - numerical_deriv) / analytical_deriv
)

fontsize <- 1.2
plot(
  log2_dx, log2(rel_err),
  frame = FALSE, # Remove the ugly box
  col = jhu_color$spiritBlue,
  xlab = "log2 stepsize for finite difference",
  ylab = "log2 rel errors",
  xlim = rev(range(log2_dx)),
  cex.lab = fontsize,
  cex.axis = fontsize,
  type = 'l'
)

```


**Questions:** Does the relative error plot agree with what the mathematical theory predicts?
Explain the observed phenomenon.

Also, what is the slope in the "latter part" of your plotted line, roughly speaking?
Explain why we observe such a slope in the relative error.
(Hint: Consider a machine with 10 decimal digits of accuracy. What happens if we numerically differentiate $f(x) = x$ at $x = 1$ with $\Delta x = 0.1, 0.01, 0.001, \ldots$.)


*Answer*: The relative error plot partially agrees with the mathematical theory. 

The log error is $log_2(\Delta x^2) = 2 log_2(\Delta x)$, so in theory, when plotting the log relative error against the log step size, the slope is 2. 

- This is what we observe in the front part (higher step size) of the error plot. The plot shows that there is a linear relationship between log2 step size and log2 relative error with slope at 2. The smaller the log2 step size, the smaller the log2 relative error. 

- But the latter part of the plot does not agree with the mathematical theory. The log2 relative error increases when log2 step size decreases. The slope in the latter part of the plotted line is roughly $-1$, and is much less smooth than the earlier part. 
  - The linear increase in (log2) relative error with smaller (log2) step size is due to finite precision of the machine representation. When decreasing the step size, the denominator $2\Delta x$ divide by 2. In theory, $f(x + \Delta x) - f(x-\Delta x)$ in the numerator should also decrease just like in the front part of the plot. However, due to finite precision error, the the numerator $f(x + \Delta x) - f(x-\Delta x)$ gets so small that it can be not represented accurately by the machine. Every time the numerator in theory should get smaller, it requires another fraction bit outside the 52 bits to be represented. The numerator then gets rounded up or down, losing precision. The log2 of the lost precision divided by step size will give us roughly $-1$, which explains a linear increase in log2 relative error against decrease in log step size
  - The unstable slope might be because the error can get either rounded up or down, so whether the estimated value is smaller or larger than the actual value is not stable, causing unstable slope

## Part B: Extention to multivariate functions

As mentioned in the lecture, the finite difference method extends to a function with multivariate input $f: \mathbb{R}^d \to \mathbb{R}$ and can be used to approximate the gradient $\nabla f(\boldsymbol{x})$.
The extension relies on the fact $f(\boldsymbol{x} + \Delta x \, \boldsymbol{e}_i) = f(\boldsymbol{x}) + \Delta x \, \partial_i f(\boldsymbol{x}) + \ldots$ for $i = 1, \ldots, d$.

Using the starter code provided below, implement a function to approximate $\nabla f(\boldsymbol{x})$ via the centered difference method.
To test this function, take as an example a quadratic function $f(\boldsymbol{x}) = - \boldsymbol{x}^\intercal \boldsymbol{\Phi} \boldsymbol{x} / 2$ with positive definite $\boldsymbol{\Phi}$, which can be thought of as a log-likelihood function for $\boldsymbol{x} \sim \mathscr{N}(\boldsymbol{0}, \boldsymbol{\Phi})$ up to an additive constant.
First show that its gradient is given by $\nabla f(\boldsymbol{x}) = - \boldsymbol{\Phi} \boldsymbol{x}$; more generally, the gradient of a quadratic function $g(\boldsymbol{x}) = \boldsymbol{x}^\intercal \boldsymbol{A} \boldsymbol{x}$ is given by $\nabla g(\boldsymbol{x} = (\boldsymbol{A} + \boldsymbol{A}^\intercal) \boldsymbol{x}$ for any square matrix $\boldsymbol{A}$.
Then compare a numerically approximated gradient to one computed via an analytical formula.

```{r}
approx_grad <- function(func, x, dx = .Machine$double.eps^(1/3)) {
  numerical_grad <- rep(0, length(x))
  # Fill in
  for (i in seq_along(x)) {
    zero_vector <- rep(0, length(x))
    zero_vector[i] <- dx
    numerical_grad[i] <- (func(x + zero_vector) - func(x - zero_vector)) / (2*dx)
  }
  return(numerical_grad)
}

set.seed(410)
n_param <- 4
X <- matrix(rnorm(2 * n_param^2), nrow = 2 * n_param, ncol = n_param)
Sigma_inv <- t(X) %*% X
  
#' Calculate log-density of centered Gaussian up to an additive factor
gaussian_logp <- function(x, Sigma_inv) {
  logp <- - .5 * t(x) %*% Sigma_inv %*% x
  return(logp)
}

gaussian_grad <- function(x, Sigma_inv) {
  #grad <- rep(0, length(x)) # Replace with an actual formula
  grad <- -1 * Sigma_inv %*% x
  return(grad)
}

x <- c(3, 1, 4, 1)

analytical_grad <- gaussian_grad(x, Sigma_inv)
numerical_grad <- approx_grad(function(x) gaussian_logp(x, Sigma_inv), x)
testthat::expect_true(are_all_close(
  analytical_grad, numerical_grad, abs_tol = Inf, rel_tol = 1e-3
))
```


## Things to commit/submit

For Part A, completed code and answers to the questions.
For Part B, derivation of the gradient and completed code that passes the test.

# Exercise 4: Continuing development of `hiperglm` --- finding MLE via `stats::optim()` and testing it

We will continue the development of the `hiperglm` package.

As discussed, our plan is to first implement an MLE finder via a pre-packaged general-purpose BFGS optimizer, against which we can then use to benchmark our custom implementation of iteratively reweighted least squares.
The idea is that using the pre-packaged function is less error-prone, especially since the gradient function is easy to unit-test.

To be extra safe, we are going to test the MLE finder via BFGS against an even simpler algorithm: the least square for a linear model via the analytical expression $\hat{\boldsymbol{\beta}}_\textrm{mle} = (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \boldsymbol{y}$.
Complete the following steps toward this goal.
Remember that you can use [`devtools::load_all()`](https://r-pkgs.org/code.html#sec-code-load-all) to help your [package development process](https://r-pkgs.org/code.html#constant-health-checks).

0.  Make and check out (`git branch` and `git checkout`) a development branch (call it `develop`) where you will commit the work below.
1.  Set up the use of `testthat`.
2.  Copy the helper functions `are_all_close()` and `simulate_data()` from the instructor's repo.
3.  Test the `are_all_close()` function. Write a test to cover at least three cases: the function 1) correctly returns `TRUE`, 2) correctly returns `FALSE` because the relative error is above `rel_tol`, and 3) correctly returns `FALSE` because the absolute error is above `abs_tol`. Make sure that the logic of the test is clear and easy to understand --- tests are there to clarify and validate the code's behavior; unreadable tests defeat the purpose. <br> *Remark:* This is just an exercise, but in general there is nothing strange about testing a function you use to test another function. It is perfectly reasonable to consider testing any functions complicated enough to potentially contain insiduous bugs.
4.  Set up the (currently failing) test to compare the MLE estimated via pseudo-inverse and via BFGS.
5.  Implement the MLE finder via peudo-inverse. Please, pretty please, do not calculate an inverse of a matrix. (Otherwise, it's fine to use `solve()` instead of `chol()`.)
6.  Implement the MLE finder via BFGS:
    a.  Implement functions to calculate the log-likelihood and gradient under a linear model. This function technically requires an estimate of the residual variance, but its value does not affect the MLE for the coefficients and hence is irrelevant for our purpose. You can just specify it as an optional parameter e.g. as `function(..., noise_var = 1)`.
    b.  Test your gradient calculation by comparing it against a numerical one. For this purpose, use the function from Exercise 2, Part B to approximate the gradient of a given function via finite difference. <br> *Remark:* Technically, this test checks only for consistency between the log-likelihood and gradient function. It is possible to have both functions wrong but consistent with each other so that the test passes. This test nonetheless can catch many bugs and, since the gradient calculation tends to be more error prone, it's good to make sure there is no mistake there.
    c.  Plug in the now tested-and-true log-likelihood and gradient functions to `stats::optim()` to find MLE.
7.  Run `devtools::test()` and check that the then-failing test from Step 2 now passes.
8.  Open a pull request from your development branch to the main/master one.

## Things to commit/submit

Link to the pull request: https://github.com/yanxiliu230/hiperglm/pull/2

