---
title: 'Homework: Numerical linear algebra'
output:
  html_notebook:
    code_folding: none
    highlight: textmate
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: {
    Macros: {
      bA: "{\\boldsymbol{A}}",
      bx: "{\\boldsymbol{x}}",
      bb: "{\\boldsymbol{b}}"
    }
  }
});
</script>


```{r setup, include=FALSE}
source(file.path("..", "R", "util.R"))

required_packages <- c("microbenchmark")
install_and_load_packages(required_packages)
```


# Exercise 1: Comparing different numerical linear algebra algorithms for solving linear systems

In this exercise, we consider the problem of solving a linear system $\bA \bx = \bb$ for $\bx$.
We compare the three methods we learned in the class: LU, Cholesky, and QR decompositions.
(Of course, LU applies to more general systems and QR can be used to solve least squares, but here we focus on positive definite systems.)

## Part A: Racing for solution &mdash; speed comparison 

We first compare their computational speed. 
Fill in the code below and `bench::mark()` the three algorithms.

**Questions:**
What are relative speeds among the algorithms?
Do relative speeds agree with what you expect from the complexity analyses?
If not (quite), why might that be?

```{r, eval=FALSE}
# Import the `rand_positive_def_matrix` function
source(file.path("R", "random_matrix.R"))

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
b <- rnorm(mat_size)

solve_via_cholesky <- function(A, b) {
  # Fill in
}

bench::mark(
  solve(A, b)
)
bench::mark(
  solve_via_cholesky(A, b)
)
bench::mark(
  # Solve via QR decomposition
)
```

## Part B: Competition in terms of numerical accuracy/stability

We now compare the three methods in terms of numerical accuracy/stability.
To this end, we set up the following simulation study. 
We first generate a "ground truth" solution vector $\bx_0$.
We then compute an "numerical-error-free" $\bb = \bA \bx_0$ by carrying out the matrix-vector multiplication using the `long double` type, which (on most hardware and compilers) provides [additional 12 bits of precision](https://en.wikipedia.org/wiki/Extended_precision#x86_extended_precision_format).
Of course, the vector $\bb$ computed as such still suffers from numerical errors, but the idea is that the numerical errors from this high-precision matrix-vector multiplication is much smaller than the errors caused by numerically solving $\bA \bx = \bb$ for $\bx$.
We can thus assess the accuracy of the three solvers by comparing the numerically computed $\bx$ to the ground truth $\bx_0$.

### Task &#x2F00;

First compare the outputs of matrix-vector multiplication $\bx \to \bA \bx$ using `double` and `long double` using the provided Rcpp functions.
We will discuss more about Rcpp in lecture (if we have not yet); you can for now just use the provided functions without worrying about how they work.

**Questions:**
What is the relative difference in $\ell^2$-norm? 
How about the coordinate-wise relative differences?
Are the observed magnitudes of the differences what you'd expect?

```{r, eval=FALSE}
Rcpp::sourceCpp(file.path("src", "matvec_double.cpp"))
Rcpp::sourceCpp(file.path("src", "matvec_ldouble.cpp"))

set.seed(1918)

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
x <- rnorm(mat_size)

# Fill in

```

### Task &#x2F06;

Now randomly generate $\bA$ so that its condition number is $10^6$.
Then solve a positive-definite system $\bA \bx = \bb$ for $\bx$ using the three algorithms and compare their outputs to the ground truth $\bx_0$.

**Questions:**
Which algorithm appears to be more accurate than the others? 
Visually demonstrate your answer.

```{r, eval=FALSE}
set.seed(1918)
cond_num <- 1e6

# Larger matrices could incur substantial computational time under base R BLAS
mat_size <- 1024L 

A <- rand_positive_def_matrix(mat_size, cond_num)
x <- rnorm(mat_size)

# Fill in

```

### Task &#x4E09;

In Task &#x2F06;, we compared the three algorithms in one randomly generated example.
Now we consider a more systematic (though hardly comprehensive) comparison via repeated simulations.
We also vary the condition number of $\bA$ and assess whether the results would hold across varying degrees of ill-conditioning.

**Questions/To-do's:**

* Using the starter code provided, calculate various summary measures of the numerical errors.
* Integrate into the provided code one another (or more, if you like) meaningful metric(s) of your choice to summarize the numerical error.
* Visually explore how the three algorithms compare with each other in their accuracy. See if different error metrics tell different stories; they might or might not.
* Vary the condition number in the range $10^6 \sim 10^{12}$, e.g. by trying $10^6$, $10^9$, and $10^{12}$.
* Do you see any patterns in the numerical errors across the three algorithms, metrics, and/or condition numbers? Show some plots to support your conclusion.

**Note:** 
The QR solver will throw an error when the system is ill-condition enough that the numerical solution might not be very accurate. 
To force it to return the solution in any case, set `tol = .Machine$double.eps`.

```{r, eval=FALSE}
# Utility functions for bookkeeping simulation results.
source(file.path("R", "num_linalg_sim_study_helper.R"))

n_sim <- 32L
mat_size <- 512L
cond_num <- 1e6
metrics <- c("norm", "median", "five_percentile", "ninety_five_percentile")
  # TODO: add another metric and modify the helper script accordingly

rel_error_list <- lapply(
  c("lu", "chol", "qr"), 
  function(method) pre_allocate_error_list(n_sim, metrics)
)

for (sim_index in 1:n_sim) {
  A <- rand_positive_def_matrix(mat_size, cond_num)
  x <- rnorm(mat_size) 
  b <- matvec_ldouble(A, x)
  x_approx <- list( 
    # Fill in
  )
  for (method in c("lu", "chol", "qr")) {
    rel_error <- lapply(
      metrics, 
      function (metric) calc_rel_error(x, x_approx[[method]], metric)
    )
    names(rel_error) <- metrics
    for (metric in names(rel_error)) {
      rel_error_list[[method]][[metric]][sim_index] <- rel_error[[metric]]
    }
  }
}

# TODO: visually compare errors

```


<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: {
    Macros: {
      X: "{\\boldsymbol{X}}",
      y: "{\\boldsymbol{y}}",
      bm: ["{\\boldsymbol #1}", 1]
    }
  }
});
</script>

# Exercise 2: Continuing development of `hiperglm` &mdash; improving numerical stability of Newton's method via QR-based solver

From the previous assignments, you should now have a `hiperglm` package that implement an MLE finder via Newton's method. 
We have so far focused on getting the functionality correct, and not so much on numerical stability and computational speed.
In this assignment, you improve the algorithm's numerical stability by turning Newton's method into the mathematically equivalent iteratively re-weighted least squares (IWLS) based on QR decomposition.
(And, in next assignment, you will improve the computational speed by interfacing the package with RcppEigen.)

Much of this and next assignments is about enhancing performance of the package's existing features, rather than adding new ones.
This is one setting in which the formal automated testing really proves its value. 
By running the existing tests after each change, you can be confident that you did not break anything by changing the internal implementations that do not affect the interface.
And, if you did change functions' interfaces, the tests help you remember where you have to make corresponding changes to ensure correct overall functionality.

Before you start this assignment, remember to first address all the feedback from the previous assignment and merge your previous work to the main branch. 
Then create a new `irls` branch from there, check it out, and commit all your work for this assignment there.
Finally, open a pull request into your main branch after completing the assignment and request a review from the TA.

1. Change your code for fitting linear models to solve the least squares problem using QR decomposition instead of LU/Cholesky.
Make sure to run the existing tests to make sure the change did not break anything.
2. Turn, if you have not already done so, each iteration of Newton's method into an isolated function.
In other words, write a function named something like `take_one_newton_step(...)` and use it within the for-loop.
This in particular facilitate unit-testing one step of Newton's method.
3. Viewing the vector $\boldsymbol{H}_f^{-1} \nabla f$ in the Newton update formula as the solution of a weighted least squares problem, implement the solver for it based on QR decomposition.
To do so, observe that minimizing $\| \bm{W}^{1/2} (\X \bm{\beta} - \y) \|^2$ is equivalent to minimizing $\| \tilde{\X} \bm{\beta} - \tilde{\y} \|^2$ with the modified input $\tilde{\X}$ and $\tilde{\y}$. 
Use base R linear algebra functions for now.
Add an option to the function `take_one_newton_step(...)` (or however else you decided to call it) to switch between the QR- and LU/Cholesky-based solvers, with QR being default.
Test that the two solvers yield the same result up to finite precision errors.
