---
title: 'Homework: Numerical linear algebra'
output:
  html_notebook:
    code_folding: none
    highlight: textmate
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: {
    Macros: {
      bA: "{\\boldsymbol{A}}",
      bx: "{\\boldsymbol{x}}",
      bb: "{\\boldsymbol{b}}"
    }
  }
});
</script>


```{r setup, include=FALSE}
source(file.path("..", "R", "util.R"))

required_packages <- c("microbenchmark")
install_and_load_packages(required_packages)
```


# Exercise 1: Comparing different numerical linear algebra algorithms for solving linear systems

In this exercise, we consider the problem of solving a linear system $\bA \bx = \bb$ for $\bx$.
We compare the three methods we learned in the class: LU, Cholesky, and QR decompositions.
(Of course, LU applies to more general systems and QR can be used to solve least squares, but here we focus on positive definite systems.)

## Part A: Racing for solution &mdash; speed comparison 

We first compare their computational speed. 
Fill in the code below and `bench::mark()` the three algorithms.

**Questions:**
What are relative speeds among the algorithms?
Do relative speeds agree with what you expect from the complexity analyses?
If not (quite), why might that be?

**Note:**
I misspoke about the R's built-in `chol()` function during the lecture:
when applied to a positive-definite `A`, the function actually returns an _upper-triangular_ matrix `R` such that `t(R) %*% R == A`.

```{r, eval=FALSE}
# Import the `rand_positive_def_matrix` function
source(file.path("R", "random_matrix.R"))

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
b <- rnorm(mat_size)

solve_via_cholesky <- function(A, b) {
  # Fill in
}

bench::mark(
  solve(A, b)
)
bench::mark(
  solve_via_cholesky(A, b)
)
bench::mark(
  # Solve via QR decomposition
)
```

## Part B: Competition in terms of numerical accuracy/stability

We now compare the three methods in terms of numerical accuracy/stability.
To this end, we set up the following simulation study. 
We first generate a "ground truth" solution vector $\bx_0$.
We then compute an "numerical-error-free" $\bb = \bA \bx_0$ by carrying out the matrix-vector multiplication using the `long double` type, which (on most hardware and compilers) provides [additional 12 bits of precision](https://en.wikipedia.org/wiki/Extended_precision#x86_extended_precision_format).
Of course, the vector $\bb$ computed as such still suffers from numerical errors, but the idea is that the numerical errors from this high-precision matrix-vector multiplication is much smaller than the errors caused by numerically solving $\bA \bx = \bb$ for $\bx$.
We can thus assess the accuracy of the three solvers by comparing the numerically computed $\bx$ to the ground truth $\bx_0$.

### Task &#x2F00;

First compare the outputs of matrix-vector multiplication $\bx \to \bA \bx$ using `double` and `long double` using the provided Rcpp functions.

**Questions:**
What is the relative difference in $\ell^2$-norm? 
How about the coordinate-wise relative differences?
Are the observed magnitudes of the differences what you'd expect?

```{r, eval=FALSE}
Rcpp::sourceCpp(file.path("src", "matvec_double.cpp"))
Rcpp::sourceCpp(file.path("src", "matvec_ldouble.cpp"))

set.seed(1918)

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
x <- rnorm(mat_size)

# Fill in

```

### Task &#x2F06;

Now randomly generate $\bA$ so that its condition number is $10^6$.
Then solve a positive-definite system $\bA \bx = \bb$ for $\bx$ using the three algorithms and compare their outputs to the ground truth $\bx_0$.

**Questions:**
Which algorithm appears to be more accurate than the others? 
Visually demonstrate your answer.

```{r, eval=FALSE}
set.seed(1918)
cond_num <- 1e6

# Larger matrices could incur substantial computational time under base R BLAS
mat_size <- 1024L 

A <- rand_positive_def_matrix(mat_size, cond_num)
x <- rnorm(mat_size)

# Fill in

```

### Task &#x4E09;

In Task &#x2F06;, we compared the three algorithms in one randomly generated example.
Now we consider a more systematic (though hardly comprehensive) comparison via repeated simulations.
We also vary the condition number of $\bA$ and assess whether the results would hold across varying degrees of ill-conditioning.

**Questions/To-do's:**

* Using the starter code provided, calculate various summary measures of the numerical errors.
* Integrate into the provided code one another (or more, if you like) meaningful metric(s) of your choice to summarize the numerical error.
* Visually explore how the three algorithms compare with each other in their accuracy. See if different error metrics tell different stories; they might or might not.
* Vary the condition number in the range $10^6 \sim 10^{12}$, e.g. by trying $10^6$, $10^9$, and $10^{12}$.
* Do you see any patterns in the numerical errors across the three algorithms, metrics, and/or condition numbers? Show some plots to support your conclusion.

**Note:** 
The QR solver will throw an error when the system is ill-condition enough that the numerical solution might not be very accurate. 
To force it to return the solution in any case, set `tol = .Machine$double.eps`.

```{r, eval=FALSE}
# Utility functions for bookkeeping simulation results.
source(file.path("R", "num_linalg_sim_study_helper.R"))

n_sim <- 32L
mat_size <- 512L
cond_num <- 1e6
metrics <- c("norm", "median", "five_percentile", "ninety_five_percentile")
  # TODO: add another metric and modify the helper script accordingly

rel_error_list <- lapply(
  c("lu", "chol", "qr"), 
  function(method) pre_allocate_error_list(n_sim, metrics)
)

for (sim_index in 1:n_sim) {
  A <- rand_positive_def_matrix(mat_size, cond_num)
  x <- rnorm(mat_size) 
  b <- matvec_ldouble(A, x)
  x_approx <- list( 
    # Fill in
  )
  for (method in c("lu", "chol", "qr")) {
    rel_error <- lapply(
      metrics, 
      function (metric) calc_rel_error(x, x_approx[[method]], metric)
    )
    names(rel_error) <- metrics
    for (metric in names(rel_error)) {
      rel_error_list[[method]][[metric]][sim_index] <- rel_error[[metric]]
    }
  }
}

# TODO: visually compare errors

```
