\documentclass{article}


\title{One reason to version control source files only}

\begin{document}

\maketitle

This is the first sentence.

This is Sentence 2.
This is Sentence 3.
This is Sentence 4.
This is Sentence 5.
This is Sentence 6.
This is Sentence 7.
This is Sentence 8.
This is Sentence 9.
This is Sentence 10.
This is Sentence 11.
This is Sentence 12.
This is Sentence 13.
This is Sentence 14.
This is Sentence 15.
This is Sentence 16.
This is Sentence 17.
This is Sentence 18.
This is Sentence 19.
This is Sentence 20.
This is Sentence 21.
This is Sentence 22.
This is Sentence 23.
This is Sentence 24.
This is Sentence 25.
This is Sentence 26.
This is Sentence 27.
This is Sentence 28.
This is Sentence 29.
This is Sentence 30.
This is Sentence 31.
This is Sentence 32.
This is Sentence 33.
This is Sentence 34.
This is Sentence 35.
This is Sentence 36.
This is Sentence 37.
This is Sentence 38.
This is Sentence 39.
This is Sentence 40.
This is Sentence 41.
This is Sentence 42.
This is Sentence 43.
This is Sentence 44.
This is Sentence 45.
This is Sentence 46.
This is Sentence 47.
This is Sentence 48.
This is Sentence 49.
This is Sentence 50.
This is Sentence 51.
This is Sentence 52.
This is Sentence 53.
This is Sentence 54.
This is Sentence 55.
This is Sentence 56.
This is Sentence 57.
This is Sentence 58.
This is Sentence 59.
This is Sentence 60.
This is Sentence 61.
This is Sentence 62.
This is Sentence 63.
This is Sentence 64.
This is Sentence 65.
This is Sentence 66.
This is Sentence 67.
This is Sentence 68.
This is Sentence 69.
This is Sentence 70.
This is Sentence 71.
This is Sentence 72.
This is Sentence 73.
This is Sentence 74.
This is Sentence 75.
This is Sentence 76.
This is Sentence 77.
This is Sentence 78.
This is Sentence 79.
This is Sentence 80.
This is Sentence 81.
This is Sentence 82.
This is Sentence 83.
This is Sentence 84.
This is Sentence 85.
This is Sentence 86.
This is Sentence 87.
This is Sentence 88.
This is Sentence 89.
This is Sentence 90.
This is Sentence 91.
This is Sentence 92.
This is Sentence 93.
This is Sentence 94.
This is Sentence 95.
This is Sentence 96.
This is Sentence 97.
This is Sentence 98.
This is Sentence 99.
This is Sentence 100.
This is Sentence 101.
This is Sentence 102.
This is Sentence 103.
This is Sentence 104.
This is Sentence 105.
This is Sentence 106.
This is Sentence 107.
This is Sentence 108.
This is Sentence 109.
This is Sentence 110.
This is Sentence 111.
This is Sentence 112.
This is Sentence 113.
This is Sentence 114.
This is Sentence 115.
This is Sentence 116.
This is Sentence 117.
This is Sentence 118.
This is Sentence 119.
This is Sentence 120.
This is Sentence 121.
This is Sentence 122.
This is Sentence 123.
This is Sentence 124.
\end{document}